---
title: "scf: An R package for analysis of Survey of Consumer Finances public-use microdata"
tags:
  - R
  - survey methodology
  - household finance
  - Survey of Consumer Finances
  - multiple imputation
  - complex surveys
authors:
  - name: Joseph N. Cohen
    orcid: 0000-0002-6197-4453
    affiliation: 1
affiliations:
  - name: Queens College, City University of New York, Queens, New York, USA
    index: 1
date: 9 May 2026
bibliography: paper.bib
---

# Summary

The *Survey of Consumer Finances* (SCF) is a triennial national survey of U.S. households' income, assets, debts, financial behavior, and household organization [@Fed2023]. The data are used in academic research, Congressional research reports, and the production of official statistics.

The R package `scf` [@Cohen2025] provides a unified framework for acquiring, preparing, analyzing, and visualizing SCF public-use microdata. It encodes standard practices for handling the survey’s design, automates implicate-level estimation and pooling, and reduces implementation complexity for common analytical tasks.

@Cohen2026 offers a more detailed description of the package design and validates its ability to produce official statistics.

# Statement of need

The valid analysis of SCF microdata requires a multi-step workflow that combines replicate-weighted survey estimation with the handling of multiply imputed datasets. Analysts must construct replicate-weighted survey designs for each implicate, estimate quantities separately across implicates, and combine results using appropriate pooling rules. 

In R, this workflow is typically implemented using Thomas Lumley's [-@Lumley2026] `survey` package along with imputed data handling tools his `mitools` [@Lumley2019] or custom scripts. These tools are flexible and methodologically appropriate but require users to manually coordinate design specification, implicate iteration, and pooling. As a result, even simple analyses involve substantial custom code, and correct implementation depends on careful adherence to SCF-specific conventions. Existing reference workflows, such as those provided by Anthony Damico [-@Damico2026], demonstrate correct implementation but require dozens of lines of code that can slow or discourage non-specialists' engagement of the data.

The `scf` package addresses this gap by formalizing SCF-specific best practices in a standardized functional interface. It provides a unified data object that encodes the replicate-weighted design across implicates, along with a set of functions that automate estimation and pooling. The package provides a consistent grammar and output format that reduces the amount of custom code required for common tasks. It reproduces official SCF statistics and reduces typical workflows from dozens of lines of custom code to a small number of standardized function calls [see @Cohen2026].

# State of the field

No dedicated R package for SCF analysis existed on CRAN prior to `scf`. 
As described above, existing approaches rely on general-purpose tools 
(`survey`, `mitools`) or reference documentation (`asdfree`) rather than 
a purpose-built functional interface. Contributing SCF-specific logic to 
`survey` would be out of scope for a general-purpose package; extending 
`asdfree` would not produce the modular, testable, CRAN-distributed 
infrastructure that enables reproducible workflows across projects and 
users. `scf` fills this gap.

# Research impact statement

The package has been validated against published Federal Reserve 
statistics; @Cohen2026 documents this validation in detail. The package 
vignettes include replications of results from two published empirical 
papers, demonstrating that `scf` produces estimates consistent with 
prior peer-reviewed work. The package is distributed through CRAN and 
targets a research community for whom the SCF is a primary data source.

# Software design

The package is designed around three principles: rigor, accessibility, and transparency. It encodes SCF-specific survey and imputation workflows in a structured object while preserving access to implicate-level survey designs for advanced use.

## Core object

The central object in the package is `scf_mi_survey`, a structured container representing one SCF wave. It stores the five imputed data sets with an associated `survey::svrepdesign` object, along with supplemental metadata. 

## Estimation workflow

Most `scf` functions follow the same internal pattern:

1. construct or retrieve the implicate-specific survey designs 
2. apply the corresponding estimation routine to each implicate
3. combine results into a pooled estimate and associated uncertainty measure
4. store the combined results and results from individual implicates in an R list

For model-based and many descriptive quantities, pooling uses Rubin-style combining rules [@Rubin1987]. For some statistics such as percentiles and proportions, the package uses appropriate implicate-aggregation strategies implemented per official documentation. The result is a consistent user interface across common SCF tasks without requiring users to manage the imputation and replicate-weight logic manually.

## Functional coverage

The package supports end-to-end SCF workflows through functions for:

- data acquisition and loading (`scf_download()`, `scf_load()`, `scf_design()`)
- data wrangling (`scf_update()`, `scf_subset()`)
- descriptive estimation (`scf_mean()`, `scf_median()`, `scf_percentile()`, `scf_freq()`, `scf_xtab()`, `scf_corr()`, `scf_percentile_sum()`)
- inference (`scf_ttest()`, `scf_prop_test()`)
- modeling (`scf_ols()`, `scf_logit()`, `scf_glm()`, `scf_quantreg()`)
- results extraction and formatting (`scf_MIcombine()`, `scf_implicates()`, `scf_regtable()`)
- visualization (`scf_plot_dbar()`, `scf_plot_cbar()`, `scf_plot_bbar()`, `scf_plot_dist()`, `scf_plot_hist()`, `scf_plot_hex()`, `scf_plot_smooth()`, `scf_theme()`)

These functions reduce repetition in common SCF workflows while remaining compatible with Thomas Lumley's [-@Lumley2026] `survey` ecosystem.

# Example

A typical workflow is concise:

```{r}
library(scf)

scf_download(2022)
scf2022 <- scf_load(2022)

scf_mean(scf2022, ~income)
```

Users can also transform variables and fit models within the same object framework:

```{r}
scf2022 <- scf_update(scf2022, senior = age >= 65)

fit <- scf_ols(scf2022, networth ~ senior)
scf_regtable(fit, digits = 3, output = "console")
```

This workflow delegates design specification, implicate iteration, and pooling to the package while leaving the substantive modeling decisions to the analyst.

Visualization capabilities also available:

```{r}

# Specify education as an ordered factor with meaningful labels
scf2022 <- scf_update(scf2022, edclF = factor(edcl, 
                                              levels = 1:4, 
                                              labels = c("<HS", "HS", 
                                                         "Associates", 
                                                         "Bachelor+"),
                                              ordered = T))

# Plot
scf_plot_cbar(scf2022, ~networth, ~edclF)

```

# Availability

`scf` is implemented entirely in R. It builds primarily on the `survey` package for complex-survey estimation and uses established CRAN infrastructure for data handling and graphics. 

The package is available on CRAN:

```{r}
install.packages("scf")
```

Source code, examples, and documentation are available at: https://github.com/jncohen/scf. The package is distributed under the MIT License.

# Acknowledgements

The package builds on the `survey` package and on publicly available SCF documentation and example workflows. The author received no external funding for this work.

# AI usage disclosure

Claude (Anthropic; versions 3.5–4.5), ChatGPT (OpenAI; GPT-4o), 
Gemini (Google; 2.0 Flash), and Copilot (Microsoft) were used as aids 
in coding, debugging, and drafting portions of the documentation and 
manuscript.

